---
title: "Untitled"
output: html_document
---
Install packages:
```{r}
library(pacman)
p_load(semPlot, tidyverse, lavaan, psych)
```
Load data:
```{r}
data <- read.delim("../data/Northdale survey _reverse coded_.dat")
```
Create the scales (make the variables for path analysis)
```{r}
#   Threat
    data$Thrt_Phy1<-data$Thrt_Phy1/3
    data$Thrt_Phy2<-data$Thrt_Phy2/3
    data$Thrt_Phy3<-data$Thrt_Phy3/3
    data$threat_scale<-data$Thrt_Phy1+data$Thrt_Phy2+data$Thrt_Phy3
#   Prejudice
    data$Att_IS1<-(data$Att_IS1-1)/6.75    
    data$Att_IS2<-(data$Att_IS2-1)/6.75  
    data$Att_IS4<-(data$Att_IS4-1)/6.75  
    data$prejudice_scale<-(data$Att_IS1 + data$Att_IS2 + data$Att_IS4)
#   Contact quantity
    data$Cnt_pos_B<-data$Cnt_pos_B/2
    data$Cnt_pos_IS1<-data$Cnt_pos_IS1/2
    data$contactquant_scale<-data$Cnt_pos_B+data$Cnt_pos_IS1
#   Contact_quality
    data$Cnt_Qul_IS1<-data$Cnt_Qul_IS1/3
    data$Cnt_Qul_IS2<-data$Cnt_Qul_IS2/3
    data$Cnt_Qul_IS3<-data$Cnt_Qul_IS3/3
    data$contactqual_scale<-data$Cnt_Qul_IS1+data$Cnt_Qul_IS2+data$Cnt_Qul_IS3
#   Empathy
    data$Empath1<-data$Empath1/3    
    data$Empath2<-data$Empath2/3
    data$Empath3<-data$Empath3/3
    data$empathy_scale<-(data$Empath1 + data$Empath2 + data$Empath3)
```
```{r}
path.dat <- data
```
Use the .R script file (first 35 lines) and data file from the first SEM seminar to test the path model that proposes: contact quality mitigates threat, and creates empathy, and these each lead to reduced prejudice – however, contact quality does not directly affect prejudice. Inspect the residuals, and add in paths you think may be important. Use the anova command to compare nested models (i.e. models that can be turned into each other by deleting or adding paths only). Construct a diagram of the final model, and annotate it appropriately. You will need to figure out how to insert that diagram into your markdown file. Interpret your model.

Step 1: Define the path model
```{r}
pathm1 <- '
      threat_scale ~ contactqual_scale
      empathy_scale ~ contactqual_scale
      prejudice_scale ~ threat_scale + empathy_scale 

'    
prejpathfit1 <- sem(pathm1, data = path.dat)
summary(prejpathfit1, fit.measures = T, standardized = T)   
```
The model above is significantly different from the saturated model (p < 0.001, df = 2), which indicates that our model may yet be improved by adding more paths, although with only two degrees of freedom, we are limited in what we can add. THe model is also significicantly different from the null model, which means that some significant improvements have been made from the no information rate (p < 0.001, df = 6). The problem with chi square measures of fit on sample sizes above 200 is that they are often statistically significant. Other measures of fit may be better suited to our analysis of this model. Both the CFI and the TLI are very low, but these incremental fit indices are reliant on chi squared values - the problems with these have been discussed above. The AIC and BIC values are noted for comparison with alternate models. We want a RMSEA of < 0.05, but our RMSEA for the above model is quite high (RMSEA = 0.25, CI = 0.19 - 0.33). We also want our SRMR to be < 0.05, and this measure comes closer top the desired value (SRMR = 0.10). 

We see from the standardized regression co-efficients that contact quality mitigates threat (beta = -0.27, p < 0.001), and predicts empathy (beta = 0.28, p < 0.001). Prejudice is predicted by threat (beta = 0.28, p < 0.001), and is mitigated (to a lesser extent) by empathy (beta = -0.20, p < 0.001).

Variances are all positive, and standard errors are all acceptably low.

Next, we consider the model visually:
```{r}
semPaths(prejpathfit1, whatLabels = "std", residuals = TRUE, nCharNodes = 15, sizeMan = 15, sizeMan2 = 5, shapeMan = "rectangle", edge.label.cex = .9, layout = "circle")
```

Next, we consider the residuals and modification indices:
```{r}
resid(prejpathfit1 , type = "normalized")
modificationindices(prejpathfit1, sort = T)

```
The covariance between threat and prejudice is overestimated (predicted as being stronger than the empirical data suggests), and the covariance between empathy and prejucdice is underestimated (predicted as being weaker than the empirical data suggests). The covariance between contact quality & threat, and contact quality & empathy (respectively) is accurately predicted by the model (no difference between observed and predicted values).
The modification indices would usually be examined to determine whether there are correlations between the error terms in the scale, but in this model we only work with scale scores and not with the manifest variables themselves, so this is not possible.

We now consider whether additionally predicting threat from contact quality could improve our model

Model with prejucice also directly predicted by contact quality:
```{r}
pathm2 <- '
      threat_scale ~ contactqual_scale
      empathy_scale ~ contactqual_scale
      prejudice_scale ~ threat_scale + empathy_scale + contactqual_scale

'    
prejpathfit2 <- sem(pathm2, data = path.dat)
summary(prejpathfit2, fit.measures = T, standardized = T)
```
We note the same issues with the chi squared values as mentioned in the first model. The CFI and TLI are much better than the first model, but still don't indicate a good fit (and are still subject to their reliance on the abovementioned chi-squared values). AIC and BIC are noted for comparison between model 1 and model 2 (see next chunk). RMSEA has gotten better (RMSEA = 0.21, CI = 0.12 - 0.31, as compared to model 1: RMSEA = 0.25, CI = 0.19 - 0.33), but still doesn't indicate a good fit (p < 0.05). SRMR has also improved (SRMR = 0.06, as compared to model 1: SRMR = 0.10), and is approaching a value that indicates good fit (p < 0.05)

The strongest predictor of prejudice is contact quality. As contact quality increases, so prejudice decreases (beta = -0.30, p < 0.001). The addition of this direct path decreases the predictive value of the mediator variables: threat (model 1 beta = 0.28, p < 0.001; model 2 beta = 0.19, p = 0.001), and empathy (model 1 beta = -0.20, p < 0.001; model 2 beta = -0.11, p = 0.06). Note that the predictive value of empathy decreases to non-significance when prejudice is predicted directly from contact quality. This suggests that prejudice is better predicted by contact quality than by the two mediating variables. 
VAriances are all positive and standard errors are all reasonably small.

We compare the AIC and BIC values for the two models to see which model best presdicts prejudice
Model 1 predicts prejudice from contact quality through two mediator variables: empathy and threat, however contact quality is not used to directly predict prejudice
Model 2 also predicts prejudice from contact quality through two mediator variables: empathy and threat, but it also predicts prejudice through contact quality directly
```{r}
anova(prejpathfit1, prejpathfit2)
```
The chi squared diffrence test indicates that the second model (which additionally predicts prejudice from contact quality directly) is significantly better than the first model (which only predicts prejudice from contact quality through the two mediator variables). We note that the second model is more complex, but only slightly (it has 1 degree of freedom less than the first model)

From the trends we noticed in the fit measures and coefficients, we decide to proceed with the second model
First we consider model 2 visually
```{r}
semPaths(prejpathfit2, whatLabels = "std", residuals = TRUE, nCharNodes = 15, sizeMan = 15, sizeMan2 = 5, shapeMan = "rectangle", edge.label.cex = .9, layout = "circle")
```
Next we inspect the residuals for model 2:
```{r}
resid(prejpathfit2 , type = "normalized")
```
We see that there is no difference in the observed and predicted covariances between contact quality & threat, contact quality & empathy, and contact quality & prejudice respectively. We also note that the difference between the observed and expected values for threat & prejudice have dropped (model 1: -0.65, model 2: -0.34), as has the difference for empathy & prejudice (model 1: 0.91, model 2: 0.61). This indictes that model 2 is better able to predict the observed data than model 1 could.

Next we look at the linear model output for the second model
We keep only the complete cases (to make things comparable)
```{r}
path.dat %<>% select(contactqual_scale, prejudice_scale, threat_scale, empathy_scale) %>% 
  na.omit()
# First compute first step regressions
lm_model_pathm2_A1 <- lm(threat_scale ~ contactqual_scale, data = path.dat)
summary(lm_model_pathm2_A1)
lm_model_pathm2_A2 <- lm(empathy_scale ~ contactqual_scale, data = path.dat)
summary(lm_model_pathm2_A2)
# Now model prejudice
lm_model_pathm2_B <- lm(prejudice_scale ~ threat_scale + empathy_scale + contactqual_scale, data = path.dat)
summary(lm_model_pathm2_B, standardized = T)
```
The regression analyses indicate:
1: Contact quality significantly lowers the perception of threat (beta = -0.25, p < 0.001)
2: Contact quality significantly increases empathy (beta = 0.26, p < 0.001)
3: Contact quality is the most significant variable that lowers prejudice (beta = -0.32, p < 0.001), followed by a low perception of threat (beta = 0.21, p = 0.001). Empathy becomes an insignificant variable in lowering prejudice (beta = 0.12, p = 0.07) when contact quality is added as a direct predictor of prejudice.

----------------------------
Test the measurement model that is implicit in 1 above. Note that the notation =~ indicates a loading in Lavaan (i.e. a path relating an indicator/manifest variable to a latent variable). What do you conclude? Check the residuals and see if this changes your mind. Construct a diagram that reports your measurement model, you only need to show latent and manifest variables

First, we get the data in the correct format:
```{r}
data2 <- read.delim("../data/Northdale survey _reverse coded_.dat")
```
Convert the manifest variables
```{r}
#   Threat
    data2$Thrt_Phy1<-data2$Thrt_Phy1/3
    data2$Thrt_Phy2<-data2$Thrt_Phy2/3
    data2$Thrt_Phy3<-data2$Thrt_Phy3/3
#   Prejudice
    data2$Att_IS1<-(data2$Att_IS1-1)/6.75    
    data2$Att_IS2<-(data2$Att_IS2-1)/6.75  
    data2$Att_IS4<-(data2$Att_IS4-1)/6.75  
#   Contact quantity
    data2$Cnt_pos_B<-data2$Cnt_pos_B/2
    data2$Cnt_pos_IS1<-data2$Cnt_pos_IS1/2
#   Contact_quality
    data2$Cnt_Qul_IS1<-data2$Cnt_Qul_IS1/3
    data2$Cnt_Qul_IS2<-data2$Cnt_Qul_IS2/3
    data2$Cnt_Qul_IS3<-data2$Cnt_Qul_IS3/3
#   Empathy
    data2$Empath1<-data2$Empath1/3    
    data2$Empath2<-data2$Empath2/3
    data2$Empath3<-data2$Empath3/3
```
```{r}
path2.dat <- data2
```
Next, we set up the measurement model and structural model implicit in the analyses above
```{r}
#   SEM of model2
    pathm2full <- '
            # MEASUREMENT MODEL
              threat_scale       =~ Thrt_Phy1 + Thrt_Phy2 + Thrt_Phy3
              contactqual_scale  =~ Cnt_Qul_IS1 + Cnt_Qul_IS2 + Cnt_Qul_IS3 
              prejudice_scale    =~ Att_IS1 + Att_IS2 + Att_IS4
              empathy_scale      =~ Empath1 + Empath2 + Empath3
      
            # STRUCTURAL MODEL
              threat_scale ~ contactqual_scale
              empathy_scale ~ contactqual_scale
              prejudice_scale ~ threat_scale + empathy_scale + contactqual_scale
                  '
```
```{r}
prejpathfit2_full <- sem(pathm2full, data = path2.dat)
    summary(prejpathfit2_full , fit.measures = T, standardized = T)
```
The model 2 above ("model 2 full" - which includes both the measurement and structural components of the model) is much better than the previous model ("model 2"), which only included the structural model (i.e.: only latent variables and not manifest variables).
Firstly the degrees of freedom have increased from 1 to 49, which will influence almost all of the fit measures for the model.
The model (model 2 full) is no longer significantly differnt from the saturated model (p = 0.17, df = 49), despite the large sample size. This indicates that, while our model is simpler than the model with all possible paths, it does not have a significantly worse predictive value. This model is still significantly differnt from the null model (p < 0.001, df = 66), indicating that, despite its reduced complexity, it is significantly better than the no-information rate.
CFI and TLI values indicate the relative position of our model between the null model and the saturated model. They should approach 1, but any value > 0.95 is considered a good fit. Both of these values indicate that our model is a good fit (CFI = 1.00, TLI = 0.99).
AIC and BIC values have dropped drastically, indicating that the full model is an improvement of our previous model (latent variables only). It is unfortunately not possible to test the significanvce of this improvement with anova, as it is not a nested model
RMSEA should approximate 0, and should be < 0.05 to indicate a good fit. The upper end of the CI should also ideally be < 0.05. RMSEA indicates our model has good fit on both accounts (RMSEA = 0.03, CI = 0.00 - 0.05).
The SRMR is an absolute measure of fit, and is thus biased towards small sample sizes and low df. This explains why this is our only test statistic that didn't improve in the full model (but surprisingly it didn't get worse, as the number of df drastically increased). SRMR should approximate 0 (which would indicate no diffrence in the observed and predicted values), but values of < 0.05 (and some say < 0.08) indicates a good fit. The SRMR for our model indicates that it has good fit, or is atleast approaching a good fit (SRMR = 0.06)

The manifest variables all significantly load onto the latent variables.
All proposed paths are significant, except for the path between empathy and prejudice (noted in the previous model interpretation):
1. Contact quality significantly mitigates the perception of threat (beta = -0.29, p < 0.001)
2. Contact quality significantly increases empathy (beta = 0.32, p < 0.001)
3. Lowered prejudice is significantly predicted by incresed contact quality (beta = -0.36, p < 0.001), and decreased perception of threat (beta = 0.19, p = 0.003). It is not significantly predicted by empathy (beta = -0.10, p = 0.12). In the previous model (model 2 - latent variables only) we made this observation, but were unable to remove empathy, as our model would become "just identified" (df = 0, therefore no fit statistics possible). Now that we have a full model with multiple degrees of freedom, we can consider whether the model would be improved by removing empathy as a mediator variable.

The variances are all positive, and the standard errors are all reasonably small.

Next, we consider the model visually:
```{r}
semPaths(prejpathfit2_full, whatLabels = "std", residuals = TRUE, nCharNodes = 15, sizeMan = 8, shapeMan = "circle", shapeLat = "rectangle", sizeLat = 14, sizeLat2 = 5, edge.label.cex = .8, layout = "tree", rotation = 2, fixedStyle = c("red", 3), optimizeLatRes = T)
```
Next, we consider the residuals and modification indices:
```{r}
resid(prejpathfit2_full , type = "normalized")
modificationindices(prejpathfit2_full, sort = T)

```
Covariance matrix indicates the difference between the covariance values observed in the sample and the covariance values expected/predicted by the model (i.e.: the residuals of the covariance matrices). The biggest residuals are for the covariances between empathy and threat - understandable, as it wasn't included in the structural model. There are also substantial residuals for the covariances between contact quality and threat (esp for contact qual item 3 & threat item 1, and contact qual items [1, 2] & threat item 3). There are also large residuals for the covariances between threat and prejudice (esp for prej item 2 and threat item 2). As expected, the lowest  residuals are for the covariances between contact quality and prejudice, and the highest residuals (of the covariances proposed by the model), are for the covariances between prejudice and empathy (esp for empathy item 1 & prej item 1, and empathy item 3 & prej item 2). Empathy item 2 has relatively low residuals in its covariance with prejudice, and may be worth investigating as a sole measure of empathy for future predictions.

Modification indices indicate that the strongest improvementss to the model could be made by:
1. Predicting empathy from prejudice. This is not our objective, as we aim to predict prejudice. However, it suggests that emapthy is predicted from prejudice and not vice versa, which may explain why empathy is not a significant predictor of prejudice in our model
2 & 3. Predicting threat from empathy and predicting empathy from threat. This suggests that while empathy may be predictive of percieved threat (which in turn is predictive of prejudice), empathy is not directly predictive of prejudice. Threat and empathy seem to have a bi-directional relationship.
4. Allowing empathy and threat residuals to covary. Empathy clearly plays a role in this model, but not in the way the model suggests. This modification reaffirms that empathy and threat may have a bidirectinal relationship.
5. Predicting threat from prejudice. Again, this isn't our objective. But as threat already significantly predicts prejudice, this suggests a bi-directional relationship.
6-9. These 4 modifications reaffirm the relationship between threat and empathy. Either the residuals of the items on the threat and empathy scales should be allowed to covary, or items from the empathy scale are manifest variables that measure the latent variable threat.

Based on the above modification indices, threat and empathy residuals should be allowed to covary
The suggested modifications for the covarying residuals between these scale items are implemented. 
```{r}
#   SEM of model3
    pathm3full <- '
            # MEASUREMENT MODEL
              threat_scale       =~ Thrt_Phy1 + Thrt_Phy2 + Thrt_Phy3
              contactqual_scale  =~ Cnt_Qul_IS1 + Cnt_Qul_IS2 + Cnt_Qul_IS3 
              prejudice_scale    =~ Att_IS1 + Att_IS2 + Att_IS4
              empathy_scale      =~ Empath1 + Empath2 + Empath3
      
            # STRUCTURAL MODEL
              threat_scale ~ contactqual_scale
              empathy_scale ~ contactqual_scale
              prejudice_scale ~ threat_scale + empathy_scale + contactqual_scale
                
            # CORRELATED ERRORS
                  Thrt_Phy1 ~~ Empath1
                  Thrt_Phy2 ~~ Empath3
                  Thrt_Phy3 ~~ Empath3

                  '
```
```{r}
prejpathfit3_full <- sem(pathm3full, data = path2.dat)
    summary(prejpathfit3_full , fit.measures = T, standardized = T)
```
Allowing these covariances of residuals of scale items improves our model (CFI = 1.00, TLI = 1.00, RMSEA = 0.00, CI = 0.00 - 0.03, SRMR = 0.06).
Empathy still doesnt significantly predict threat, but its predictive value has improved (model 2 full: beta = -0.10, p = 0.12; model 3 full: beta = -0.11, p = 0.10)

An inspection of our covariances of residuals indicate that threat item 2 does not significantly covary with empathy item 2 or empathy item 3. We can attempt to remove these non-significant covariations:


```{r}
#   SEM of model4
    pathm4full <- '
            # MEASUREMENT MODEL
              threat_scale       =~ Thrt_Phy1 + Thrt_Phy2 + Thrt_Phy3
              contactqual_scale  =~ Cnt_Qul_IS1 + Cnt_Qul_IS2 + Cnt_Qul_IS3 
              prejudice_scale    =~ Att_IS1 + Att_IS2 + Att_IS4
              empathy_scale      =~ Empath1 + Empath2 + Empath3
      
            # STRUCTURAL MODEL
              threat_scale ~ contactqual_scale
              empathy_scale ~ contactqual_scale
              prejudice_scale ~ threat_scale + empathy_scale + contactqual_scale
                
            # CORRELATED ERRORS
                  Thrt_Phy1 ~~ Empath1

                  '
```
```{r}
prejpathfit4_full <- sem(pathm4full, data = path2.dat)
    summary(prejpathfit4_full , fit.measures = T, standardized = T)
```
Removing non-significant covariations of residuals does not improve the model. So while these covariations are not significant on their own, they do improve the model overall when allowed. We reatin model 3.

We inspect the model (model 3 full) visually:
```{r}
semPaths(prejpathfit3_full, whatLabels = "std", residuals = TRUE, nCharNodes = 15, sizeMan = 8, shapeMan = "circle", shapeLat = "rectangle", sizeLat = 14, sizeLat2 = 5, edge.label.cex = .8, layout = "tree", rotation = 2, fixedStyle = c("red", 3), optimizeLatRes = T, curve = 3, curvature = .7)
```
The three item covariances can be seen on the far right hand side.
-------------------------------------------------------------------------

By default, Lavaan tests oblique factor structures (i.e. where factors are allowed to correlate with each other). Force Lavaan to test an orthogonal factor structure for the measurement model in 2, and compare results.

```{r}
prejpathfit3_full_orth <- sem(pathm3full, data = path2.dat, orthogonal = TRUE)
    summary(prejpathfit3_full_orth , fit.measures = T, standardized = T)
```
Changing the rotation to orthogonal did not alter the model noticably.



